%% PRE %% 0 "thesis.tex"
%% PRE %% 0 "<built-in>"
%% PRE %% 0 "<command-line>"
%% PRE %% 1 "/usr/include/stdc-predef.h" 1 3 4
%% PRE %% 0 "<command-line>" 2
%% PRE %% 1 "thesis.tex"
\documentclass{report}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[backend=biber,sorting=none]{biblatex}
\usepackage{mathtools}
\usepackage{dsfont}
\usepackage{tikz}
\usetikzlibrary{topaths,calc}

\addbibresource{temp.bib}

\title{A Faster Algorithm for the No-Rainbow Problem}

\author{Edvard Thornros}

\begin{document}
\maketitle

\begin{abstract}

%% PRE %% 1 "abstract.tex" 1
No Rainbow is a constraint satisfaction problem.
The No Rainbow problem is finding a surjective node coloring of an r-regular hyper-graph.
This thesis extends on the work done by Ghazaleh Parvini and David Fernandez-Baca ~\cite{sourceNoRainbow}.
A faster algorithm using equivalence classes is suggested, evaluated and proved to be correct.

Other ideas for potential algorithms are also suggested.
%% PRE %% 22 "thesis.tex" 2

\end{abstract}

%% PRE %% 1 "intro.tex" 1
\chapter{Introduction}
The faster something can be done, the faster technology can iterate.
A part of this is finding faster algorithms for known hard problems.
One such hard problem is the no rainbow problem, which is NP-complete.
The no rainbow problems asks if there is a surjective node coloring of an $r$-regular hypergraph.
\cite{sourceNoRainbow}

Rephrased, the no rainbow problem asks if we can assign each node a color.
In a graph with $r$ nodes per edge -- each edge connects $r$ different nodes.
All $r$ colors should be represented in the entire graph.
No edge should connect all $r$ colors.
A node coloring which satisfies these constraints is a solution to the no rainbow problem.

An edge which connects all $r$ colors is called a rainbow edge -- hence the name of the problem.
The no rainbow problem is a variant of the constraint satisfaction problem.

The no rainbow problem is related to the phylogenetic decisiveness problem. The phylogenetic decisiveness problem asks if a potential ancestral tree for a species of animals could possibly be correct, given only parts of the information about the ancestry. \cite{sourcePhylogeneticDecisiveness}
% TODO: Read https:
% TODO: https:

\section{Research questions}
\begin{enumerate}
  \item Is there a *faster deterministic algorithm that solves the no rainbow problem than the one suggested by Ghazaleh Parvini and David Fernandez-Baca \cite{sourceNoRainbow}?
  \item Is there a *faster randomized algorithm that solves the no rainbow problem than the one suggested by Ghazaleh Parvini and David Fernandez-Baca \cite{sourceNoRainbow}?
  % \item How can a solution to the no rainbow problem be found, understood, modeled and presented?
  % \item What insights into algorithms are found by studying the no rainbow problem?
  \item How *fast is the *fastest possible algorithm for solving the no rainbow problem?
\end{enumerate}
*speed is defined by asymptotic time complexity.

\section{Aim}
The aim of this study is to find an asymptotically faster algorithm that solves
the no rainbow problem.

\section{Delimitations}
The scope of this study is to investigate if there is an asymptotically faster
algorithm that solves the no rainbow problem.
%% PRE %% 26 "thesis.tex" 2
%% PRE %% 1 "theory.tex" 1
\chapter{Background}
The no rainbow problem is quite complex and can be understood in a lot of different ways.
To understand the problem as broadly as possible a few different areas of discrete mathematics are needed.

\section{Equivalence classes and equivalence relations}
Equivalence classes are used to group things that are equal in some sense.
This requires we have a set of objects and an binary operator we can use to check if they are equal.
In this general case we use $\sim$ as the operator.

$$
  a \sim b \implies \textit{$a$ is related to $b$ under the equivalence relation ($\sim$)}
$$

A common way of defining equivalence relations is to map the objects to their image and then use equality.
More generally we can define it using:
$$
  a \sim b \iff f(a) = f(b)
$$
Where $f(a)$ is the image of $a$.

For example:
$$
  a \sim b \iff (a \bmod 3) = (b \bmod 3) : \quad a, b \in \mathds{Z}
$$

Here we send each object to their image -- which is their remained after division by 3 -- and then compare them.
Here we get 3 equivalence classes $\{[0], [1], [2]\}$. All integers are mapped into exactly one of these 3 equivalence class.

The representative of a class is the element which has itself as image. In other words, $f(a) = a \iff a$ is a representative.
\cite[Section 7.3]{sourceArmen} \cite[Section 1.2]{sourceAATA}

\section{Graphs}
Graphs consist of a set of edges and a set of nodes.
Nodes are usually presented as circles with edges connecting the circles as lines, as presented in \ref{figGraphExample}.
Usually an edge is some kind of relationship.
A graph is a very powerful and general tool for modeling many problems.

\begin{center}
\begin{figure}[h]
\centering
\begin{tikzpicture}[main/.style = {draw, circle}]
    \node[main] (A) {};
    \node[main] (B) [left of=A] {};
    \node[main] (C) [below of=A] {};
    \node[main] (D) [left of=C] {};
    \node[main] (E) [left of=D] {};
    \node[main] (F) [below of=E] {};

    \draw (A) -- (B);
    \draw (A) -- (C);
    \draw (B) -- (D);
    \draw (B) -- (E);
    \draw (F) -- (E);
\end{tikzpicture}
  \caption{An example of a graph.}
  \label{figGraphExample}
\end{figure}
\end{center}

\cite[Chapter 1]{sourceDiestel}
\cite[Chapter 1]{sourceGWA}
\cite[Section 9.1]{sourceArmen}

% Does this add anything?

\section{Hyper Graphs}
A hyper graph is similar to a normal graph.
But in a hyper graph edges connect multiple nodes, usually more than 2.
A hyper edge which is 2-regular is a "normal" graph.
The "normal" graph can be seen as a special case of the hyper graph.

Another way of phrasing it is that an edge is a set of nodes instead of a pair.

A hyper graph is the kind of graph the no rainbow problem is interesting for.

\begin{center}
\begin{figure}[h]
\centering
\begin{tikzpicture}[main/.style = {draw, circle}]
    \node[main] (a) at (0,2) {a};
    \node[main] (b) at (2,2) {b};
    \node[main] (c) at (4,2) {c};
    \node[main] (d) at (0,4) {d};
    \node[main] (e) at (2,4) {e};
    \node[main] (f) at (4,4) {f};

    \begin{scope}[fill opacity=0.5]
    \filldraw[fill=yellow!70]
        plot [smooth cycle, tension=1.5]
        coordinates {
          ($(a)+(-0.8,0)$)
          ($(b)+(0,0.8)$)
          ($(c)+(0.8,0)$)
          ($(b)+(0,-0.8)$)
        };

    \filldraw[fill=blue!70]
        plot [smooth cycle, tension=0.8]
        coordinates {
          ($(c)+(0,-0.8)$)
          ($(f)+(0.8,0.8)$)
          ($(e)+(-0.4,0.4)$)
        };

    \filldraw[fill=red!70]
        plot [smooth cycle, tension=1.5]
        coordinates {
          ($(d)+(-0.8,0)$)
          ($(e)+(0,0.8)$)
          ($(f)+(0.8,0)$)
          ($(e)+(0,-0.8)$)
        };

    \filldraw[fill=green!70]
        plot [smooth cycle, tension=1.0]
        coordinates {
          ($(c)+(-0.4,-0.4)$)
          ($(c)+(0.4,0.4)$)
          ($(e)+(0,0.8)$)
          ($(a)+(-0.4,0.4)$)
          ($(a)+(0.4,-0.4)$)
          ($(e)+(0,-0.8)$)
        };

    \end{scope};

    \node[main] (a) at (0,2) {a};
    \node[main] (b) at (2,2) {b};
    \node[main] (c) at (4,2) {c};
    \node[main] (d) at (0,4) {d};
    \node[main] (e) at (2,4) {e};
    \node[main] (f) at (4,4) {f};
\end{tikzpicture}
  \caption{An example of a hyper graph.}
  \label{figHyperGraphExample}
\end{figure}
\end{center}

% Show the other way of rendering them

The hyper graph shown in figure \ref{figHyperGraphExample} is $3$-regular.
A hyper graph is $r$-regular if and only if all edges connect $r$-different nodes.
\cite{sourceHyper}

\subsection{Node coloring as related by the no rainbow problem}
A node coloring is a mapping from nodes to colors.
There are usually constraints on what colors can be part of an edge.
The color of a node is denoted as $c(a)$, where $a$ is the node and $c$ is the coloring.
Since a coloring is a mapping it can be thought of as a function that takes you from a node to the color of the node.
A color is usually a letter or a number.

The constraints added in the no rainbow problem are
\begin{enumerate}
  \item No edge can have all r-colors -- no rainbow edges
  \item All colors have to be represented in the graph -- the coloring is surjective
\end{enumerate}

An example of a valid coloring in the no rainbow problem for the graph presented above is $c(d)=0, c(b)=1, c(*)=2$, where $*$ denotes all other nodes.
There are a lot more valid colorings, though this coloring is particularly simple.
The validity of a coloring is dictated by the edges added in the graph.
A hyper-graph with the maximum number of edges (called a complete graph) has no valid no-rainbow coloring.
A hyper-graph without any edges is trivial to find a no-rainbow coloring for.
The cases in between are the main topic of this thesis.

\cite{sourceHyper}
% Is a complete r-regular subgraph enough to disprove there is a no-rainbow coloring?

\section{The existing algorithms}
TBA

\section{Algorithm basics}
TBA
%% PRE %% 27 "thesis.tex" 2

\printbibliography

\end{document}
